---
title: "Case Study: Bike Sharing"
output: html_document

---


## Install & loading the required packages
#====================================================

```{r}
install.packages("tidyverse")
install.packages("lubridate")
install.packages("skimr")
install.packages("janitor")
install.packages("dplyr")
library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(skimr) #get summary data
library(janitor)#compare dataframe
library(dplyr)
```


# STEP 1: COLLECT DATA
#====================================================

```{r}
Trips_Apr20 <- read_csv('202004-divvy-tripdata.csv')
Trips_May20 <- read_csv('202005-divvy-tripdata.csv')
Trips_June20 <- read_csv('202006-divvy-tripdata.csv')
Trips_July20 <- read_csv('202007-divvy-tripdata.csv')
Trips_Aug20 <- read_csv('202008-divvy-tripdata.csv')
Trips_Sep20 <- read_csv('202009-divvy-tripdata.csv')
Trips_Oct20 <- read_csv('202010-divvy-tripdata.csv')
Trips_Nov20 <- read_csv('202011-divvy-tripdata.csv')
Trips_Dec20 <- read_csv('202012-divvy-tripdata.csv')
Trips_Jan21 <- read_csv('202101-divvy-tripdata.csv')
Trips_Feb21 <- read_csv('202102-divvy-tripdata.csv')
Trips_Mar21 <- read_csv('202103-divvy-tripdata.csv')
Trips_Apr21 <- read_csv('202004-divvy-tripdata.csv')
```


# STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE
#====================================================
# Compare column names each of the files
# While the names don't have to be in the same order, they DO need to match perfectly before we can use a command to join them into one file


```{r}
colnames(Trips_Apr20)
colnames(Trips_May20)
colnames(Trips_June20)
colnames(Trips_July20)
colnames(Trips_Aug20)
colnames(Trips_Sep20)
colnames(Trips_Oct20)
colnames(Trips_Nov20)
colnames(Trips_Dec20)
colnames(Trips_Jan21)
colnames(Trips_Feb21)
colnames(Trips_Mar21)
colnames(Trips_Apr21)
```


# Inspect the dataframes and look for inconguencies
#====================================================

```{r}
str(Trips_Apr20)
str(Trips_May20)
str(Trips_June20)
str(Trips_July20)
str(Trips_Aug20)
str(Trips_Sep20)
str(Trips_Oct20)
str(Trips_Nov20)
str(Trips_Dec20)
str(Trips_Jan21)
str(Trips_Feb21)
str(Trips_Mar21)
str(Trips_Apr21)
```

# we can compare column datatype across all dataframe by using compare_df_cols when we have large dataset, that would be more easy
```{r}

compare_df_cols(Trips_Apr20, Trips_May20, Trips_June20, Trips_July20, Trips_Aug20, Trips_Sep20, Trips_Oct20, Trips_Nov20, Trips_Dec20, 
Trips_Jan21, Trips_Feb21, Trips_Mar21, Trips_Apr21, return = "mismatch")
```

# Convert end_station_id and start_station_id to character so that they can stack correctly

```{r}
Trips_Apr20 <- mutate(Trips_Apr20, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_May20 <- mutate(Trips_May20, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_June20 <- mutate(Trips_June20, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_July20 <- mutate(Trips_July20, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_Aug20 <- mutate(Trips_Aug20, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_Sep20 <- mutate(Trips_Sep20, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_Oct20 <- mutate(Trips_Oct20, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_Nov20 <- mutate(Trips_Nov20, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
Trips_Apr21 <- mutate(Trips_Apr21, end_station_id = as.character(end_station_id), start_station_id = as.character(start_station_id))
```

# double check column datatype across all dataframe

```{r}
compare_df_cols(Trips_Apr20, Trips_May20, Trips_June20, Trips_July20, Trips_Aug20, Trips_Sep20, Trips_Oct20, Trips_Nov20, Trips_Dec20, 
Trips_Jan21, Trips_Feb21, Trips_Mar21, Trips_Apr21, return = "mismatch")
```


# Stack individual data frames into one big data frame

```{r}
all_trips <- bind_rows(Trips_Apr20, Trips_May20, Trips_June20, Trips_July20, Trips_Aug20, Trips_Sep20, Trips_Oct20, Trips_Nov20, Trips_Dec20, 
Trips_Jan21, Trips_Feb21, Trips_Mar21, Trips_Apr21)
```

# Remove unused column
```{r}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng))
```

# Rename Columns
```{r}
all_trips <- all_trips %>%  rename(trip_id= ride_id ,ride_type = rideable_type 
                  ,start_time = started_at,end_time =ended_at
                  ,from_station_name = start_station_name
                  ,from_station_id = start_station_id 
                  ,to_station_name = end_station_name 
                  ,to_station_id = end_station_id
                  ,usertype = member_casual)
```



# STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS
#======================================================
# Inspect the new table that has been created
```{r}
colnames(all_trips)  #List of column names
dim(all_trips)  #Dimensions of the data frame?
head(all_trips) #See the first 6 rows of data frame.
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics
skim(all_trips) #get summary of data, check missing data 
```


# Add columns that list the date, month, day, and year of each ride
# This will allow us to aggregate ride data for each month, day, or year ... before completing these operations we could only aggregate at the ride level
```{r}
all_trips$date <- as.Date(all_trips$start_time) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```


# Add a "ride_length" calculation to all_trips (in seconds)
```{r}
all_trips$ride_length <- difftime(all_trips$end_time,all_trips$start_time)
```

# Convert "ride_length" from Factor to numeric so we can run calculations on the data
```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

# Remove "bad" data
# The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative
```{r}
skim(all_trips$ride_length)
all_trips_v2 <- all_trips[!(all_trips$ride_length<0),]
skim(all_trips_v2)
```


# STEP 4: CONDUCT DESCRIPTIVE ANALYSIS
#=====================================
# Descriptive analysis on ride_length (all figures in seconds)
```{r}
summary(all_trips_v2$ride_length)
```

# Export to CSV file for further analysis
```{r}
write.csv(all_trips_v2, "data.csv")
```


